---
title: "Visualizing Chicago Public Transit Ridership"
date: '2019-11-23'
tags:
- R Markdown
- spatial maps
- segmentation
categories:
- Data Visualization
- Spatial Analysis
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# For all practicle purposes this should always be FALSE 
knitr::opts_chunk$set(widgetframe_self_contained = FALSE) # default = FALSE

# For all practicle purposes this should always be TRUE 
knitr::opts_chunk$set(widgetframe_isolate_widgets = TRUE) # default = TRUE
```

In this post, I used Open data from Chicago Transit Authority (CTA) to segment routes based on ridership, and developed a visualization dashboard to identify stops with unbalanced boardings and alightings and stops with the most ridership. Get the raw data [here](https://raw.githubusercontent.com/enjieli/CAT-data-demo/master/app/CTA_-_Ridership_-_Avg._Weekday_Bus_Stop_Boardings_in_October_2012.csv).


### load package
```{r echo = TRUE, warning=FALSE, message=FALSE}
library(knitr)
library(tidyverse)
library(stringr)
library(sf)
library(kableExtra)
library(RColorBrewer)
library(viridis)
library(leaflet)
library(ggmap)
library(widgetframe)
```



```{r echo = TRUE, warning=FALSE, message=FALSE}

raw_df<- read_csv("CTA_-_Ridership_-_Avg._Weekday_Bus_Stop_Boardings_in_October_2012.csv")
```

```{r echo = TRUE, warning=FALSE, message=FALSE}
#clean raw_df 
raw_df<- 
  raw_df %>%
  mutate( location= str_replace_all(location, "[()]", "")) %>%
  separate(location, c("lat", "lon"), sep = ",") %>%
  mutate( lat =as.numeric(lat)) %>%
  mutate( lon= as.numeric(lon))
```


</div>
<div style="margin-top:50px;">
##### Table 1 Routes with most stops
<div style="margin-bottom:20px;">
</div>


```{r echo = TRUE, warning=FALSE, message=FALSE,results = 'asis'}

#What is the route with the most stops?
routes <- 
  raw_df %>%
  separate_rows(routes) %>%
  group_by(routes) %>%
  summarise(Number_of_stops=n_distinct(stop_id)) %>%
  arrange(desc(Number_of_stops))

kable(routes[1:10, ],  booktabs = T) %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "left")


```


</div>
<div style="margin-top:100px;">
##### Table 2 Stops with the most routes
<div style="margin-bottom:20px;">
</div>

```{r echo = TRUE, warning=FALSE, message=FALSE}
#What is the stop with the most routes?
stops <- 
  raw_df %>%
  mutate(number_of_routes = str_count(routes, "\\d+")) %>%
  arrange(desc(number_of_routes)) %>%
  select(stop_id,on_street, cross_street, routes, number_of_routes)

kable(stops[1:10, ], booktabs = T) %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "left")
```


</div>
<div style="margin-top:100px;">
##### Figure 1 Stop Density Map
<div style="margin-bottom:20px;">
</div>

```{r chicago_basemap, echo = TRUE, warning=FALSE, message=FALSE}
#get chicago basemap 
chi_bb <- c(left = -87.88430,
            bottom = 41.64416,
            right = -87.52570,
            top = 42.06470)

chicago_stamen <- get_stamenmap(bbox = chi_bb, zoom = 11, maptype = 'toner-lite')

# calculate stop density
df_sp <- 
  st_as_sf(raw_df, coords = c("lon", "lat"), crs = 4326) %>%
  select(-c('month_beginning','daytype'))

#plot them together
stops_density_map <- 
  ggmap(chicago_stamen)+ 
  stat_density_2d(data = raw_df,
                  aes(x = lon, y = lat, fill = stat(level)),
                  alpha = .4,
                  bins = 5,
                  geom = "polygon") +
  scale_fill_gradientn('number of stops',colors = brewer.pal(5, "YlOrRd")) +
  geom_sf(data = df_sp, inherit.aes = FALSE,color='black',
          size = 0.2, alpha=0.1,show.legend = 'point') 

stops_density_map

```


</div>
<div style="margin-top:100px;">
##### Figure 2 Map of stops with no ridership
<div style="margin-bottom:20px;">
</div>

```{r stop_route_long, echo = TRUE,  warning=FALSE, message=FALSE}
#tranform to long dataframe
stops_route <- 
  raw_df %>%
  mutate(number_of_routes = str_count(routes, "\\d+")) %>%
  separate_rows(routes) %>%
  mutate( boardings = boardings/number_of_routes) %>%
  mutate(alightings =  alightings/ number_of_routes)  %>%
  select( -c(number_of_routes,daytype,month_beginning)) %>%
  gather("boardings", "alightings", key = type, value = ridership)

#make it sf object 
stops_route_sp <- 
  st_as_sf(stops_route, coords = c("lon", "lat"), crs = 4326) 

#stops with no boardings and alightings
#transform stops_no_ridership to spatial dataframe

stops_no_ridership_sp <-
  stops_route_sp %>%
  filter (ridership==0) %>%
  group_by(stop_id, routes) %>%
  filter(n()>1)

#n_distinct(stops_no_ridership_sp$stop_id)
#n_distinct(stops_no_ridership_sp$routes)

#set color scheme
pal <- colorFactor(palette = 'Spectral', domain =stops_no_ridership_sp$routes)

#plot
p1<-leaflet(stops_no_ridership_sp) %>%
  setView(lng=-87.705, lat=41.85443, zoom = 10) %>% 
  addProviderTiles(providers$Stamen.Toner,group="Stamen Toner") %>%
  addCircleMarkers( color = ~pal(routes),
                        radius = ~ ifelse(ridership <=100, 6,
                                           ifelse((ridership <=1000),10,15)),
                        stroke = FALSE, 
                        fillOpacity = 0.8,
                        popup = ~paste("Stop_id:", stop_id, 
                          "<br>", "Route:", routes, 
                          "<br>", "Ridership:", ridership)) %>%
      addLegend("topright", pal = pal, opacity = 1,
                values = ~routes, 
                title = "Routes")
frameWidget(p1, width='100%')

```



</div>
<div style="margin-top:100px;">
##### Table 3  Top 10 most trafficked routes  
<div style="margin-bottom:20px;">
</div>

```{r most_trafficked, echo = TRUE, warning=FALSE, message=FALSE}
#routes by boarding and alighting ridership 
ridership_by_routes<- 
  stops_route %>%
  filter(!is.na(routes)) %>%
  group_by(routes, type) %>%
  summarise(n=sum(ridership)) %>%
  spread(type, n) %>%
  mutate(total =boardings+ alightings) %>%
  arrange(desc(total))

kable(ridership_by_routes[1:10, ], booktabs = T) %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "left")
```


</div>
<div style="margin-top:100px;">
##### Figure 3 Stops segmentation based on Ridership
<div style="margin-bottom:20px;">
</div>


```{r kmeans df setup,  echo = TRUE,warning=FALSE, message=FALSE}
#reshape df with only lat, lon, boardings, and alights columns
#scale the dataset
km_df_scale<- 
  raw_df%>%
  select(stop_id,boardings,alightings) %>%
  column_to_rownames(var="stop_id") %>%
  scale()
```

```{r determin_k,  echo = TRUE, warning=FALSE, message=FALSE}
library(factoextra)
set.seed(123)
# function to compute total within-cluster sum of square 
wss <- function(k) {
  kmeans(km_df_scale, k,iter.max = 100, nstart = 25 )$tot.withinss
}

# Compute and plot wss for k = 1 to k = 15
k.values <- 1:15

# extract wss for 2-15 clusters
wss_values <- map_dbl(k.values, wss)

#plot optimal K
plot(k.values, wss_values,
       type="b", pch = 19, frame = FALSE, 
       xlab="Number of clusters K",
       ylab="Total within-clusters sum of squares")
```     

```{r kmeans,  echo = TRUE, warning=FALSE, message=FALSE}
set.seed(123)

#kmeans with 3 clusters
km.res <- kmeans(km_df_scale, 3, iter.max = 100, nstart = 25)

#join with raw df
dd <- cbind(raw_df, cluster = km.res$cluster)

#plot
dd %>%
  ggplot(aes(x=boardings, y= alightings, color= factor(cluster))) +
  geom_point() +
  geom_abline(aes(intercept = 0 , slope= 1 ), linetype= "dashed") + 
  scale_x_continuous(trans = 'log') +
  scale_y_continuous(trans = 'log') +
  scale_size(range = c(0.5, 5))  +
  scale_color_brewer("", palette = "Set1", 
                     labels = c("Stops with little ridership", 
                                "Stops with most ridership", 
                                "Stops with medium ridership")) +
  theme_bw() +
  xlab("Boardings") +
  ylab("Alightings") 
```

```{r }
#visualize on a map
m = leaflet() %>%
    setView(lng=-87.705, lat=41.85443, zoom = 10) %>% 
    addProviderTiles(providers$Stamen.Toner,group="Stamen Toner") 

# subset into clusters 
data_filterlist = list(cluster_1 = subset(dd, cluster == 1),
                cluster_2 = subset(dd, cluster == 2),
                cluster_3 = subset(dd, cluster == 3))

# Remember we also had these groups associated with each variable? Let's put them in a list too:
layerlist = c("Stops with little ridership", "Stops with most ridership", "Stops with medium ridership")

# We can keep that same color variable:
colorFactors = colorFactor(c('red', 'blue', 'seagreen'), domain=dd$cluster)

# Now we have our loop - each time through the loop, it is adding our markers to the map object:

for (i in 1:length(data_filterlist)){
  m = addCircleMarkers(m, 
                       lng=data_filterlist[[i]]$lon, 
                       lat=data_filterlist[[i]]$lat, 
                       radius=1,
                       popup= paste( "Stop_id:",data_filterlist[[i]]$stop_id,  
                                     "<br>", "Routes:", data_filterlist[[i]]$routes,
                                     "<br>", "Boardings:",data_filterlist[[i]]$boardings,
                                     "<br>", "Alightings:",data_filterlist[[i]]$alightings),
                       stroke = FALSE, fillOpacity = 0.75,
                       color = colorFactors(data_filterlist[[i]]$cluster),
                       group = layerlist[i]
  )

}


p2<- addLayersControl(m, overlayGroups = layerlist, options = layersControlOptions(collapsed = FALSE))

frameWidget(p2, width='100%')

```




